Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-07-04 survey/survey.tex
all of 4.3 now written.

Here we will focus instead of the syntactic interpretation as
program transformers. This results in a function
\begin{flalign*}
\mathit{eval}_1\,:\,\ \left\{ t_1\, t_2\,:\,U\right\}
  \left\{ c_1\, c_2\,:\, t_1\,\leftrightarrow\,t_2\right\}
  (ce\,\,c_1 \Leftrightarrow\,c_2) \rightarrow (t_1\,\leftrightarrow\,t_2) \\
\end{flalign*}
This function is ``deeply dependent'': given the type of the
rewrite $ce$ to apply, both the input $c_1$ and output $c_2$ are
almost entirely determined!  Let us take for example the
second combinator in Fig.~\ref{figi}:
\[
  {(c_1 \otimes (c_2 \oplus c_3)) \odot \distl \Leftrightarrow \distl \odot ((c_1 \otimes c_2) \oplus (c_1 \otimes c_3))}
\]
\noindent which we can name \AgdaInductiveConstructor{distl⇔l}.
Interpreting this as a rewrite from the program on the left to the
one on the right requires ``pattern matching'' on the left
structure which contains $3$ arbitrary combinators, from which
we can \emph{reconstruct} the program on the right. Rewrites
such as \AgdaInductiveConstructor{distl⇔l} are one-step rewrites,
in the same way that \distl is a constant of the base term language
of $\Pi$.  There is one additional wrinkle. There is naturally
an opposite combinator, which interprets the above from right
to left; let us call it \AgdaInductiveConstructor{distl⇔r}. It
would appear to require \emph{non-linear pattern-matching}
since the right-hand-side contains $c_1$ twice. That is however not
the case! The definition of \AgdaInductiveConstructor{distl⇔r} has
$5$ implicit arguments, $3$ of which are $c_1, c_2$, and $c_3$,
which then completely force the ``shape'' of the overall pattern.
Thus the mere mention of \AgdaInductiveConstructor{distl⇔r} is
enough to resolve the apparent use of a non-linear pattern.
This is why $\mathit{eval}_1$ was called ``deeply dependent''
above: once the name of the combinator is given, the rest
follows.

>>>>

\newcommand{\evalone}{\ensuremath{\mathit{eval}_1}}

Here we will focus instead of the syntactic interpretation as
program transformers. This results in a function
\begin{flalign*}
\mathit{eval}_1\,:\,\ \left\{ t_1\, t_2\,:\,U\right\}
  \left\{ c_1\, c_2\,:\, t_1\,\leftrightarrow\,t_2\right\}
  (ce\,\,c_1 \Leftrightarrow\,c_2) \rightarrow (t_1\,\leftrightarrow\,t_2) \\
\end{flalign*}
This function is ``deeply dependent'': given the type of the
rewrite $ce$ to apply, both the input $c_1$ and output $c_2$ are
almost entirely determined!  Let us take for example the
second combinator in Fig.~\ref{figi}:
\[
  {(c_1 \otimes (c_2 \oplus c_3)) \odot \distl \Leftrightarrow \distl \odot ((c_1 \otimes c_2) \oplus (c_1 \otimes c_3))}
\]
\noindent which we can name \AgdaInductiveConstructor{distl⇔l}.
Interpreting this as a rewrite from the program on the left to the
one on the right requires ``pattern matching'' on the left
structure which contains $3$ arbitrary combinators, from which
we can \emph{reconstruct} the program on the right. Rewrites
such as \AgdaInductiveConstructor{distl⇔l} are one-step rewrites,
in the same way that \distl is a constant of the base term language
of $\Pi$.  There is one additional wrinkle. There is naturally
an opposite combinator, which interprets the above from right
to left; let us call it \AgdaInductiveConstructor{distl⇔r}. It
would appear to require \emph{non-linear pattern-matching}
since the right-hand-side contains $c_1$ twice. That is however not
the case! The definition of \AgdaInductiveConstructor{distl⇔r} has
$5$ implicit arguments, $3$ of which are $c_1, c_2$, and $c_3$,
which then completely force the ``shape'' of the overall pattern.
Thus the mere mention of \AgdaInductiveConstructor{distl⇔r} is
enough to resolve the apparent use of a non-linear pattern.
This is why \evalone was called ``deeply dependent''
above: once the name of the combinator is given, the rest
follows.

\newcommand{\transLR}{\AgdaInductiveConstructor{trans⇔}}
If all expressible transformations were single-step only, this
would hardly justify calling this an ``operational semantics'',
as we would hardly have a programming language. However, this level
of $\Pi$ has combinators as well: two are in
Fig.~\ref{fige} and two are in Fig.~\ref{figh}. The most interesting
one is ``sequential composition'', which is the middle one at the 
bottom of Fig.~\ref{figh}. Since $\Leftrightarrow$ represents
an equivalence, sequential composition in this context is the
same as transitivity of equivalences, as thus we have chosen to
name this \transLR. When evaluating \transLR, we could cheat:
we know that the eventual answer must be, and we could just 
return that. But this is not operational in any real sense, as
that skips over the intermediate steps. We would like to be able
to ``trace'' the rewrite. Thus the evaluation of
$\transLR r_0 r_1$ where $r_0 : c_0 \leftrightarrow c_1$ and
$r_1 : c_1 \leftrightarrow c_2$
should apply \evalone to both $r_0$ and $r_1$. Furthermore,
after applying $r_0$, we should be able to witness that the result
is indeed $c_1$, so that we may continue. This last requirement
forces us to define a new function, mutually recursively with
\evalone, for this task:
\begin{flalign*}
\mathit{exact}\,:\,\ \left\{ t_1\, t_2\,:\,U\right\}
  \left\{ c_1\, c_2\,:\, t_1\,\leftrightarrow\,t_2\right\}
  (ce\,\,c_1 \Leftrightarrow\,c_2) \rightarrow \evalone\, ce \equiv c_2 \\
\end{flalign*}
\noindent If we are careful in our construction of \evalone, the
definition of $\mathit{exact}$ is quite straightforward, i.e.
almost all cases are \AgdaInductiveConstructor{refl}.

This then lets us define the $\transLR r_0 r_1$ case properly: we
first evaluate $r_0$ and get a result combinator, witness that this result
type is indeed exactly what we expect, and proceed to evaluate $r_1$
where we specific that the $r_1$'s left-hand side must be
$\evalone r_0$; we can use \AgdaKeyword{rewrite} to make this match
$c_2$ ``on the nose'' (otherwise the call would be ill-typed).
This then forces us to use \AgdaKeyword{rewrite} also in the
implementation of the \transLR case in $\mathit{exact}$ before
the proof is indeed just \AgdaInductiveConstructor{refl}.

The other three combinators are much simpler, as simple
recursive calls are sufficient.
